---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r}
library(synergyfinder)
library(tidyverse)

`%notin%` <- Negate(`%in%`)

getwd()
load("data/mlr.master.RData")
```

```{r}
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::mutate)
```

# Functions
```{r}
# format data for synergyfinder
# data is in replicate, so calculate average response and then normalize values to the highest avg response and finally convert to percentage
conv_cytok_to_inhib <- function(data) {
  
  # determine unique concentration pairs, since results are in replicate
  temp <- data
  temp$unique_conc <- paste0(signif(temp$conc1, 4), "_", signif(temp$conc2, 4))
  temp$unique_conc <- factor(temp$unique_conc, levels = unique(temp$unique_conc))
  value <- numeric(0)
  
  # calculate mean value for each concentration pair
  for (i in levels(temp$unique_conc)) {
    
    value <- c(value, mean(temp$response_norm[temp$unique_conc == i]))
    # print(i)
    
  }
  
  # format table for synergyfinder
  temp2 <- unique(temp[, c("block_id", "assay_type", "assay_number", "drug1", "drug2", "conc1", "conc2", "conc_unit1", "conc_unit2", "analyte", "donor_1_id", "donor_2_id")])
  colnames(temp2) <- c("block_id", "assay_type", "assay_number", "drug1", "drug2", "conc1", "conc2", "conc_unit1", "conc_unit2", "analyte", "donor_1_id", "donor_2_id")
  temp2$avg_response <- value
  
  # normalize response to highest value and convert to percent
  max_total <- max(temp2$avg_response, na.rm = TRUE)
  print(max_total)
  temp2$response <- (temp2$avg_response/max_total)*100
  
  return(temp2)
}
```


# Subset columns to df of interest
```{r}
head(mlr.master)

mlr.df <- mlr.master %>% 
  ungroup() %>% 
  select(assay_type, assay_number, 
         treatment_1, concentration_1, treatment_2, concentration_2, 
         analyte, analyte_value,
         donor_1_id, donor_2_id)

dim(mlr.df)

# how many unique analyte readings from the experiments?
unique.experiments <- mlr.df %>% 
  select(assay_type, assay_number, treatment_1, treatment_2, analyte, donor_1_id, donor_2_id) %>% 
  unique() %>% 
  drop_na() # 264

```

# Normalize each donor pair to corresponding no-treatment controls
```{r}
mlr.normalized <- NULL

for(i in 1:nrow(unique.experiments)) {
  # i = 235
  unique.experiments[i, ]
  
  dataxls0 <- mlr.df %>% 
    filter(assay_number == unique.experiments$assay_number[i], 
           assay_type == unique.experiments$assay_type[i],
           treatment_1 == unique.experiments$treatment_1[i],
           treatment_2 == unique.experiments$treatment_2[i],
           analyte == unique.experiments$analyte[i], 
           donor_1_id == unique.experiments$donor_1_id[i],
           donor_2_id == unique.experiments$donor_2_id[i]
    )
  
  # format data for synergyfinder
  df0.temp <- dataxls0 %>%
    add_column(block_id = i, .before = 1) %>%
    mutate(conc_unit1 = "ug/mL") %>%
    mutate(conc_unit2 = "ug/mL")
  
  dataxls <- df0.temp %>%
    select(block_id,
           drug1 = treatment_1,
           drug2 = treatment_2,
           conc1 = concentration_1,
           conc2 = concentration_2,
           analyte,
           response = analyte_value,
           conc_unit1,
           conc_unit2,
           donor_1_id,
           donor_2_id,
           assay_type, assay_number) %>%
    drop_na()
  
  # View(dataxls)
  
  # change drug names into factors
  dataxls$drug1 <- as.factor(dataxls$drug1)
  dataxls$drug2 <- as.factor(dataxls$drug2)
  
  # normalize the data to no treatment control mean, by donor
  control <- dataxls %>% filter(conc1 == 0 & conc2 == 0) # get all control samples
  norm <- mean(control$response) # calculate mean of controls
  dataxls$response_norm <- dataxls$response - norm
  
  mlr.normalized <- rbind(mlr.normalized, dataxls)
}

# remove results where data could not be normalized
mlr.normalized <- mlr.normalized %>% 
  filter(!is.na(response_norm))


# remove experiments where readouts are too few (i.e., > 10 rows)
filter.out <- mlr.normalized %>% 
  group_by(block_id) %>% 
  summarise(num_readouts = n()) %>% 
  filter(num_readouts < 10)

mlr.normalized <- mlr.normalized %>% 
  filter(block_id %notin% filter.out$block_id)
```


# Convert cytokine values to % induction
- avg_response is mean value of replicate readouts (not yet converted to % induction)
- response is % induction values
```{r}
# hold %induction values that synergyfinder takes
df.induction <- NULL

unique.blocks <- unique(mlr.normalized$block_id) # 240

for(i in unique.blocks) {
  # average any technical replicates and convert to percent induction values
  df <- mlr.normalized %>% 
    filter(block_id == i) %>% 
    conv_cytok_to_inhib(.)
  
  df.induction <- rbind(df.induction, df)
}
# View(df.induction)

# remove results where response is lower than no-treatment
## caused when highest value is zero (from no treatment, cannot divide by zero)
df.induction %>% 
  filter(is.na(response) | is.infinite(response)) # 222

df.induction <- df.induction %>% 
  filter(!is.na(response) & !is.infinite(response)) # 9135 - 222 = 8913
```

# Run synergyfinder
```{r}
# Reshape the data to be useable by synergyfinder
res <- ReshapeData(
  data = df.induction,
  data_type = "inhibition",
  impute = TRUE,
  impute_method = NULL,
  noise = TRUE,
  seed = 1)
# 196 missing values are imputed

# Calculate synergy
res <- CalculateSynergy(
  data = res,
  method = c("ZIP", "HSA", "Bliss", "Loewe"),
  Emin = NA,
  Emax = NA,
  correct_baseline = "non")
str(res$synergy_scores)

View(res$drug_pairs)
View(res$response)
View(res$synergy_scores)
```


# Merge drug pair synergy results with df.induction
```{r}
drug.pairs <- full_join(
  df.induction %>% 
    select(block_id, assay_type, assay_number, drug1, drug2, analyte, donor_1_id, donor_2_id) %>% 
    unique(),
  res$drug_pairs %>% 
    mutate(
      ZIP_synergy_p_value = round(as.numeric(ZIP_synergy_p_value), 3),
      HSA_synergy_p_value = round(as.numeric(HSA_synergy_p_value), 3),
      Bliss_synergy_p_value = round(as.numeric(Bliss_synergy_p_value), 3),
      Loewe_synergy_p_value = round(as.numeric(Loewe_synergy_p_value), 3)
    )
)

View(drug.pairs)

write_csv(drug.pairs, file = "~/Genmab/mlr_app/data/mlr-master_drug-pairs.csv")

# filter to relevant columns for Han
drug.pairs.subset <- drug.pairs %>% 
  filter(analyte %in% c("IL-2", "IFNg", "TNF-a")) %>% 
  select(block_id, assay_type, assay_number, drug1, drug2, analyte, donor_1_id, donor_2_id, HSA_synergy) %>% 
  rename(mean_HSA_synergy = HSA_synergy)

write_csv(drug.pairs.subset, file = "~/Genmab/mlr_app/data/mlr-master_drug-pairs_subset.csv")
```


# Merge synergy scores with df.induction
```{r}
synergy.scores <- full_join(
  df.induction %>% 
    rename(response_readout = avg_response,
                response_induction = response), 
  res$synergy_scores, 
  by = c("block_id", "conc1", "conc2"))

View(synergy.scores)

write_csv(synergy.scores, file = "~/Genmab/mlr_app/data/mlr-master_synergy-scores.csv")

# filter to relevant columns for Han
synergy.scores.subset <- synergy.scores %>% 
  filter(analyte %in% c("IL-2", "IFNg", "TNF-a")) %>% 
  select(block_id, assay_type, assay_number, drug1, conc1, drug2, conc2, analyte, donor_1_id, donor_2_id, response_readout, response_induction, HSA_synergy)

write_csv(synergy.scores.subset, file = "~/Genmab/mlr_app/data/mlr-master_synergy-scores_subset.csv")
```

